Boutiques: a flexible framework to integrate command-line applications in computing platforms

### 2018-05-09 boutiquespaper.tex
Removed all footnotes, as per the editor's request

Computational platforms such as web services, portals, science
gateways, workflow engines and virtual research environments commonly
integrate third-party applications to enable various types of data
processing. Applications, however, are often manually and
repeatedly integrated in such platforms whereas automating and
sharing this effort would improve computational
reproducibility~\cite{peng2011reproducible,Stodden1240} and contribute
to Open Science. Meanwhile, container systems such as
Docker\footnote{\url{https://www.docker.com}} and
Singularity~\cite{kurtzer2017singularity} have emerged to facilitate
the sharing and migration of software by defining immutable, reusable
execution environments.

@@

\paragraph{Publisher} As \boutiques has primarily been adopted in the neuroinformatics
community, the publisher gets more information about the described application (such as author,
website, etc.), and adds an index to it
on
NeuroLinks\footnote{\url{https://brainhack101.github.io/neurolinks}},
a repository containing links to neuroinformatics resources and
tools. We intentionally opted for publishing descriptors to an index
such as NeuroLinks rather than to a common \boutiques repository so
that descriptors remain in the tool repositories and can be maintained
by the tool developers directly. We are currently extending
\texttt{bosh} to enable users to search for \boutiques tools in
NeuroLinks using the command-line or Python API. The publishing
functionality could be extended to new repositories for other domains,
such as Bioconductor for
bioinformatics~\cite{gentleman2004bioconductor}.

@@

\subsubsection{Nipype}
Nipype
(\url{http://nipype.readthedocs.io/en/latest})~\cite{gorgolewski2011nipype}
is a workflow engine widely used in neuroinformatics. Nipype workflows
can be composed from \boutiques applications using the Python API. As
an example, we implemented
NipBIDS\footnote{\url{https://github.com/big-data-lab-team/sim/tree/master/sim/other_wf_examples/nipype}},
a Nipype workflow to process BIDS datasets using BIDS apps imported as
\boutiques applications. NipBIDS iterates participant analyses on all
the subjects found in a BIDS dataset and runs a group analysis if
requested.

@@

The Virtual Imaging Platform (VIP)~\cite{GLAT-13} is a web portal for
medical simulation and image data analysis. VIP makes applications
available as services, and connects them to the biomed Virtual
Organization (VO) in the European Grid
Infrastructure\footnote{\url{https://www.egi.eu}}. The biomed VO
interconnects approximately 65 computing sites world-wide and provides
access to 130 computing clusters and 5~PB of storage.  The VIP service
is deployed at the Creatis
laboratory\footnote{\url{https://www.creatis.insa-lyon.fr}} in Lyon
and it uses the DIRAC French national
service\footnote{\url{https://dirac.france-grilles.fr}} to execute
jobs on EGI grid and cloud resources.  As of October 2017, VIP counts
more than 1,000 registered users and a growing number of available
applications.

@@

\paragraph{Anima}
\texttt{animaN4BiasCorrection}, an ITK-based bias field correction application from the
Anima\footnote{\url{https://github.com/Inria-Visages/Anima-Public/wiki}}
project was made available in VIP through \boutiques.

@@

\paragraph{GinFizz}
We integrated the
GinFizz\footnote{\url{https://github.com/thomashirsch/ginfizz}}
Nipype-based fMRI pre-processing pipeline in VIP with \boutiques.  A
few technical issues coming from the management of users in Docker
containers had to be addressed: to enable the execution in \boutiques
we had to override the permissions of files and folders in the GinFizz
container. We also had to install all the GinFizz pipeline components
in a single container while multiple ones were used by the application
initially.

@@

\paragraph{MRTrix3}
A few applications from the MRtrix3
package~\cite{tournier2012mrtrix}\footnote{\url{http://www.mrtrix.org/}} for diffusion MRI processing were also made available in VIP,
as well as a pipeline developed at Creatis, which combines MRtrix3 and
FSL applications.

@@

\paragraph{ODIN} The Odin MRI simulator~\cite{jochimsen2004odin}\footnote{\url{http://od1n.sourceforge.net}} was integrated
in VIP through \boutiques. Since Odin requires important amounts of
computing resources it is executed by VIP on the EGI grid that
currently does not support Docker. The Docker image was used just for
compilation and testing; the Odin executable was extracted from the
Docker image and the Odin wrapper script was modified accordingly.

@@

We used \boutiques to integrate 23 pipelines in the VIP platform in
the context of two challenges organized by the MICCAI conference in
2016, related to the segmentation of multiple-sclerosis lesions in MR
images (MSSEG
challenge\footnote{\url{https://portal.fli-iam.irisa.fr/msseg-challenge/overview}})
and of tumor volumes in PET images (PETSEG
challenge~\cite{hatt2018}). The
pipelines were integrated in VIP and executed on 205 subjects in a few
weeks only. Some pipelines had to be adjusted manually once integrated in
the platform for the following reasons:
\begin{enumerate}
\item A pipeline
  required a GPU, which we enabled through the
  \texttt{nvidia-docker}\footnote{\url{https://github.com/NVIDIA/nvidia-docker}}
  tool not supported in Boutiques although it could be a possible extension.
\item A pipeline required more than 10 GB of data dependencies (atlas
  data) which exceeded the maximum size allowed for Docker containers
  in our setup. We solved the issue by installing the data in a
  directory of the host server that we mounted in the container.
\item A pipeline wrote more than 10 GB of intermediate data in a
  temporary directory of the container located on a 2 GB partition. We
  solved the issue by mounting a host directory in the temporary directory. 
\end{enumerate}
Two custom properties (\texttt{vip:miccai-challenger-email}
and \texttt{vip:miccai-challenge-team-name}) were also added to the
\boutiques descriptor to help post-process results in the specific
context of MICCAI challenges.

@@

The Common Workflow Language
(CWL\footnote{\url{http://www.commonwl.org}}) is the work most closely
related to \boutiques as it provides a formal way to describe
containerized applications. In particular, CWL's Command Line Tool
Description overlaps with the \boutiques descriptor. This Section
highlights the main differences between CWL and \boutiques, based on
version 1.0 of the CWL Command Line Tool
Description\footnote{\url{http://www.commonwl.org/v1.0/CommandLineTool.html}}. According
to GitHub, CWL started 6 months before \boutiques (September 2014
vs. May 2015).

@@

Several other frameworks have been created to facilitate the
integration of command-line applications in platforms. In
neuroinformatics, many platforms define a formal interface to embed
command-line applications. Among them, the Common
Toolkit\footnote{\url{http://www.commontk.org}} interoperates with
several platforms such as 3D Slicer~\cite{pieper20043d},
NiftyView~\cite{Craddock2016}, GIMIAS~\cite{larrabide2009gimias},
MedInria~\cite{larrabide2009gimias}, MeVisLab~\cite{heckel2009object}
and MITK workbench~\cite{nolden2013medical}. The framework, however,
remains tightly bound to the Common Toolkit's C++ implementation which
limits its adoption, e.g., in web platforms.

>>>>

Computational platforms such as web services, portals, science
gateways, workflow engines and virtual research environments commonly
integrate third-party applications to enable various types of data
processing. Applications, however, are often manually and
repeatedly integrated in such platforms whereas automating and
sharing this effort would improve computational
reproducibility~\cite{peng2011reproducible,Stodden1240} and contribute
to Open Science. Meanwhile, container systems such as
Docker (\url{https://www.docker.com}) and
Singularity~\cite{kurtzer2017singularity} have emerged to facilitate
the sharing and migration of software by defining immutable, reusable
execution environments.

@@

\paragraph{Publisher} As \boutiques has primarily been adopted in the neuroinformatics
community, the publisher gets more information about the described application (such as author,
website, etc.), and adds an index to it
on
NeuroLinks (\url{https://brainhack101.github.io/neurolinks}),
a repository containing links to neuroinformatics resources and
tools. We intentionally opted for publishing descriptors to an index
such as NeuroLinks rather than to a common \boutiques repository so
that descriptors remain in the tool repositories and can be maintained
by the tool developers directly. We are currently extending
\texttt{bosh} to enable users to search for \boutiques tools in
NeuroLinks using the command-line or Python API. The publishing
functionality could be extended to new repositories for other domains,
such as Bioconductor for
bioinformatics~\cite{gentleman2004bioconductor}.

@@

\subsubsection{Nipype}
Nipype
(\url{http://nipype.readthedocs.io/en/latest})~\cite{gorgolewski2011nipype}
is a workflow engine widely used in neuroinformatics. Nipype workflows
can be composed from \boutiques applications using the Python API. As
an example, we implemented
NipBIDS (\url{https://github.com/big-data-lab-team/sim/tree/master/sim/other_wf_examples/nipype}),
a Nipype workflow to process BIDS datasets using BIDS apps imported as
\boutiques applications. NipBIDS iterates participant analyses on all
the subjects found in a BIDS dataset and runs a group analysis if
requested.

@@

The Virtual Imaging Platform (VIP)~\cite{GLAT-13} is a web portal for
medical simulation and image data analysis. VIP makes applications
available as services, and connects them to the biomed Virtual
Organization (VO) in the European Grid
Infrastructure (\url{https://www.egi.eu}). The biomed VO
interconnects approximately 65 computing sites world-wide and provides
access to 130 computing clusters and 5~PB of storage.  The VIP service
is deployed at the Creatis
laboratory in Lyon
and it uses the DIRAC French national
service (\url{https://dirac.france-grilles.fr}) to execute
jobs on EGI grid and cloud resources.  As of October 2017, VIP counts
more than 1,000 registered users and a growing number of available
applications.

@@

\paragraph{Anima}
\texttt{animaN4BiasCorrection}, an ITK-based bias field correction application from the
Anima (\url{https://github.com/Inria-Visages/Anima-Public/wiki})
project was made available in VIP through \boutiques.

@@

\paragraph{GinFizz}
We integrated the
GinFizz (\url{https://github.com/thomashirsch/ginfizz})
Nipype-based fMRI pre-processing pipeline in VIP with \boutiques.  A
few technical issues coming from the management of users in Docker
containers had to be addressed: to enable the execution in \boutiques
we had to override the permissions of files and folders in the GinFizz
container. We also had to install all the GinFizz pipeline components
in a single container while multiple ones were used by the application
initially.

@@

\paragraph{MRTrix3}
A few applications from the MRtrix3
package~\cite{tournier2012mrtrix} (\url{http://www.mrtrix.org}) for diffusion MRI processing were also made available in VIP,
as well as a pipeline developed at Creatis, which combines MRtrix3 and
FSL applications.

@@

\paragraph{ODIN} The Odin MRI simulator~\cite{jochimsen2004odin} (\url{http://od1n.sourceforge.net}) was integrated
in VIP through \boutiques. Since Odin requires important amounts of
computing resources it is executed by VIP on the EGI grid that
currently does not support Docker. The Docker image was used just for
compilation and testing; the Odin executable was extracted from the
Docker image and the Odin wrapper script was modified accordingly.

@@

We used \boutiques to integrate 23 pipelines in the VIP platform in
the context of two challenges organized by the MICCAI conference in
2016, related to the segmentation of multiple-sclerosis lesions in MR
images (MSSEG
challenge -- \url{https://portal.fli-iam.irisa.fr/msseg-challenge/overview})
and of tumor volumes in PET images (PETSEG
challenge~\cite{hatt2018}). The
pipelines were integrated in VIP and executed on 205 subjects in a few
weeks only. Some pipelines had to be adjusted manually once integrated in
the platform for the following reasons:
\begin{enumerate}
\item A pipeline
  required a GPU, which we enabled through the
  \texttt{nvidia-docker} (\url{https://github.com/NVIDIA/nvidia-docker})
  tool not supported in Boutiques although it could be a possible extension.
\item A pipeline required more than 10 GB of data dependencies (atlas
  data) which exceeded the maximum size allowed for Docker containers
  in our setup. We solved the issue by installing the data in a
  directory of the host server that we mounted in the container.
\item A pipeline wrote more than 10 GB of intermediate data in a
  temporary directory of the container located on a 2 GB partition. We
  solved the issue by mounting a host directory in the temporary directory. 
\end{enumerate}
Two custom properties (\texttt{vip:miccai-challenger-email}
and \texttt{vip:miccai-challenge-team-name}) were also added to the
\boutiques descriptor to help post-process results in the specific
context of MICCAI challenges.

@@

The Common Workflow Language
(CWL -- \url{http://www.commonwl.org}) is the work most closely
related to \boutiques as it provides a formal way to describe
containerized applications. In particular, CWL's Command Line Tool
Description overlaps with the \boutiques descriptor. This Section
highlights the main differences between CWL and \boutiques, based on
version 1.0 of the CWL Command Line Tool
Description (\url{http://www.commonwl.org/v1.0/CommandLineTool.html}). According
to GitHub, CWL started 6 months before \boutiques (September 2014
vs. May 2015).

@@

Several other frameworks have been created to facilitate the
integration of command-line applications in platforms. In
neuroinformatics, many platforms define a formal interface to embed
command-line applications. Among them, the Common
Toolkit (\url{http://www.commontk.org}) interoperates with
several platforms such as 3D Slicer~\cite{pieper20043d},
NiftyView~\cite{Craddock2016}, GIMIAS~\cite{larrabide2009gimias},
MedInria~\cite{larrabide2009gimias}, MeVisLab~\cite{heckel2009object}
and MITK workbench~\cite{nolden2013medical}. The framework, however,
remains tightly bound to the Common Toolkit's C++ implementation which
limits its adoption, e.g., in web platforms.
